---
title: "Tidy Tuesday - Wealth & Income"
description: |
  Visualizing disparities in Wealth & Income in the US over the last 40 years as
  part of [Tidy Tuesday](https://github.com/rfordatascience/tidytuesday), a weekly
  data project.
author:
  - name: Logan Lossing
    url: {}
date: 02-08-2021
output:
  distill::distill_article:
    self_contained: false
draft: false
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)

```


```{r libs_data, echo=TRUE, results=FALSE, warning=FALSE, message=FALSE}
# Libs
library(tidyverse)
library(gganimate)
library(tidytuesdayR)
library(ggtext)
library(ggdark)

# Load list of tbls
tuesdata <- tidytuesdayR::tt_load('2021-02-09')

income_tbl <- tuesdata$income_distribution
lifetime_earn_tbl <- tuesdata$lifetime_earn
```


# Income Distribution

I have been meaning to spend more time with `gganimate` and these data provided
and opportunity.

What extra value does `gganimate` versus regular time-series type plots?

The ability to draw attention.

Especially when asking the reader a question, you are providing a task to be
accomplished. By using animation, we can direct focus to the where the action is.

In this particular example, I think it is really powerful to see the proportion of 
high-income white individuals shoot up while the proportion of high-income Hispanic
and Black individuals climbs very slowly.


```{r income_distribution}

income_levels <- c(
  "Under $15,000",
  "$15,000 to $24,999",
  "$25,000 to $34,999",
  "$35,000 to $49,999",
  "$50,000 to $74,999",
  "$75,000 to $99,999",
  "$100,000 to $149,999",
  "$150,000 to $199,999",
  "$200,000 and over"
)

# Define colors for race
color_group_1 <- "#8900ff"
color_group_2 <- "#fff205"
color_group_3 <- "#ff1178"


plot <- income_tbl %>% 
  filter(race %in% c("Black Alone", "White Alone", "Hispanic (Any Race)")) %>%
  filter(year >= 1980) %>%
  filter(income_bracket %in% c('Under $15,000','$75,000 to $99,999','$200,000 and over')) %>%
  mutate(race = case_when(
    race == 'Black Alone' ~ 'Black',
    race == 'White Alone' ~ 'White',
    race == 'Hispanic (Any Race)' ~ 'Hispanic'
  )) %>%
  mutate(income_dist_percent = income_distribution / 100) %>%
  mutate(income_bracket = factor(income_bracket, levels = income_levels)) %>%
  
  ggplot(aes(x = year, y = income_dist_percent, color = race, group = race)) +
    geom_path(size = 1.2, alpha = 0.7) +
    coord_cartesian(clip = 'off') +
    scale_y_continuous(position = 'right', name = ' ', labels = scales::percent) +
    scale_color_manual(values = c(color_group_1, color_group_2, color_group_3)) +
    labs(
      title = "Is income disparity by race disappearing over time or getting worse?",
      subtitle = glue::glue("Non-White households are overrepresented in lower income brackets and are not experiencing \n  similar increases in representation in higher income brackets as White households."),
      x = NULL
    ) +
  
    transition_reveal(along = year) +
    
    facet_grid(
      rows = vars(income_bracket), 
      switch = 'y', 
      labeller = labeller(income_bracket = label_wrap_gen(width = 10))
    ) +
    
    dark_theme_minimal() +
    theme(legend.position = 'top',
          strip.text.y.left = element_text(angle = 0),
          legend.title = element_blank(),
          plot.background = element_rect(color = 'black'),
          panel.grid.major.x = element_line(color = "#44475a", size = 0.3),
          panel.grid.major.y = element_line(color = "#44475a", size = 0.3),
          panel.spacing.y = unit(2, "lines"))
  

animate(plot, height = 6, width = 9, units = "in", res = 200)
anim_save("income_disparity.gif")

```

## Income Distribution 2 - The Bug I Need to Figure Out

I wanted to make the previous plot even easier to follow by printing out the
percentage for the current year at the current end of the time series. At the
moment, I have am having an issue with this plot as the plotted text and points
from the first frame persist through the animation.

```{r income_distribution2}

income_levels <- c(
  "Under $15,000",
  "$15,000 to $24,999",
  "$25,000 to $34,999",
  "$35,000 to $49,999",
  "$50,000 to $74,999",
  "$75,000 to $99,999",
  "$100,000 to $149,999",
  "$150,000 to $199,999",
  "$200,000 and over"
)

# Define colors for race
color_group_1 <- "#8900ff"
color_group_2 <- "#fff205"
color_group_3 <- "#ff1178"


plot <- income_tbl %>% 
  filter(race %in% c("Black Alone", "White Alone", "Hispanic (Any Race)")) %>%
  filter(year >= 1980) %>%
  filter(income_bracket %in% c('Under $15,000','$75,000 to $99,999','$200,000 and over')) %>%
  mutate(race = case_when(
    race == 'Black Alone' ~ 'Black',
    race == 'White Alone' ~ 'White',
    race == 'Hispanic (Any Race)' ~ 'Hispanic'
  )) %>%
  mutate(income_dist_percent = income_distribution / 100) %>%
  mutate(income_bracket = factor(income_bracket, levels = income_levels)) %>%
  
  ggplot(aes(x = year, y = income_dist_percent, color = race, group = race)) +
    geom_path(size = 1.2, alpha = 0.7) +
    coord_cartesian(clip = 'off') +
    scale_y_continuous(position = 'right', name = ' ', labels = scales::percent) +
    scale_color_manual(values = c(color_group_1, color_group_2, color_group_3)) +
    labs(
      title = "Is income disparity by race disappearing over time or getting worse?",
      subtitle = glue::glue("Non-White households are overrepresented in lower income brackets and are not experiencing \n  similar increases in representation in higher income brackets as White households."),
      x = NULL
    ) +
    
    geom_point(size = 2) + 
    geom_text(aes(x = year + 1, label = scales::percent(income_dist_percent, accuracy = 0.1)), hjust = 0) + 
    transition_reveal(along = year) +
    
    facet_grid(
      rows = vars(income_bracket), 
      switch = 'y', 
      labeller = labeller(income_bracket = label_wrap_gen(width = 10))
    ) +
    
    dark_theme_minimal() +
    theme(legend.position = 'top',
          strip.text.y.left = element_text(angle = 0),
          legend.title = element_blank(),
          plot.background = element_rect(color = 'black'),
          panel.grid.major.x = element_line(color = "#44475a", size = 0.3),
          panel.grid.major.y = element_line(color = "#44475a", size = 0.3),
          panel.spacing.y = unit(2, "lines"))
  
animate(plot, height = 6, width = 9, units = "in", res = 200)
anim_save("income_disparity_buggy.gif")

```

# Average Lifetime Earnings


```{r avg_lifetime, fig.height = 6, fig.width = 9}

lifetime_earn_wm <- lifetime_earn_tbl %>% 
  filter(gender == 'Men', race == 'White') %>%
  pull()

color_group_1 <- "#d21404"
color_group_2 <- "#ff6961"

wealth_plot <- lifetime_earn_tbl %>%
  mutate(percent_white_male = (lifetime_earn - lifetime_earn_wm) / lifetime_earn_wm ) %>%
  mutate(diffinc_white_male = lifetime_earn - lifetime_earn_wm) %>%
  mutate(race = if_else(race == 'Hispanic any race', 'Hispanic', race)) %>%
  mutate(category = str_glue("{race} \n {gender}")) %>%
  filter(race != "White" | gender != "Men") %>%
  arrange(percent_white_male) %>%
  ggplot(aes(x = reorder(category, percent_white_male), y = percent_white_male, color = gender)) +
  expand_limits(y = 0) +
  geom_bar(stat = "identity", width = 0.01) + 
  geom_point(shape = 25) +
  scale_color_manual(values = c(color_group_1, color_group_2)) +
  labs(
    title = "Economic Disparity Charges Interest",
    subtitle = glue::glue("The average lifetime earnings in 2016 for White Men was {scales::dollar(lifetime_earn_wm)}. \n By comparison, Women and BIPOC make significantly less...but by how much?"),
    x = NULL,
    caption = "Data: Urban Institute and the US Census."
  ) + 
  guides(color = FALSE, fill = FALSE) +
  dark_theme_minimal() +
  scale_y_continuous(
    
    # first axis
    name = " ",
    labels = scales::percent,
    
    # Add a second axis
    sec.axis = sec_axis(
      ~.*lifetime_earn_wm, 
      name=" ",
      breaks = c(0, -233000, -650000, -1300000), 
      labels = c("($0)", 
                 glue::glue("Avg Cost to Raise Child \n ($233,000)"), 
                 glue::glue("Median Home Price \n Los Angeles \n ($650,000)"),
                 glue::glue("Median Home Price \n San Francisco \n ($1,300,000)"))
      )
    ) + 
  
  theme(
      axis.ticks = element_blank(),
      legend.title = element_blank(),
      plot.background = element_rect(color = 'black'),
      plot.caption = element_text(size = 8, lineheight = .9),
      plot.caption.position = "plot",
      plot.margin = margin(1, 2, 1, 1, unit = "line"),
    ) 

wealth_plot
  

```

