---
title: "SQL Basics from the Terminal"
description: |
  Learn how to create a SQL database quickly from the terminal. Useful for
  beginners who want a no nonsense bare bones method of playing with SQL databases
  or for anyone looking to create minimal examples for 
  testing and development.

preview: sql_terminal.png
author:
  - name: Logan Lossing
    url: {}
date: 01-30-2021
tags: ['SQL', 'terminal', 'database']
categories:
  - SQL
  - terminal
  - databases
output:
  distill::distill_article:
    self_contained: false
    code_folding: false
    toc: true
    toc_depth: 3
    number_sections: true
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

# Why learn SQL from the terminal?

Without an IDE to worry about or any real configuration details you can focus
on just SQL basics...with none of the headaches. Personally, I often create small
examples tables when designing a new database or trying to design a complex query.

For this quick tutorial, we will be using SQLlite. There are many flavors of 
SQL and this one is the quickest and easiest to install.

# Install SQLite

The [SQLite.org](https://www.sqlite.org/download.html) contains instructions for
installing on your operating system but for macOS, you can quickly install SQLite using homebrew `brew install sqlite`. Once installed, verify that SQLite is installed
properly using `sqlite3 --version`.


# CRUD

The core SQL database operations are known by the acronym C.R.U.D. - (C)reate, (R)ead, (U)pdate, and (D)elete so that should be a good first set of thing to learn about.

If you are not coming from a strong database background and are more comfortable
with R or python, you may first want to soon get a feel for how a table in a database is different from a tibble/data_frame. The good news is, in practice, much more is
similar than is different. However, databases offer an entire array of new possibilities
and control...enough to know in fact that managing databases can be an entire
job. 

## Create Tables

Now that we have SQLite installed, we probably want to first create some tables.

To create a database containing a table, we need only provide names for the columns of the
database and provide SQL with information about the type of data for each column.
The [SQLite documentation](https://www.sqlite.org/datatype3.html) contains extensive
information about all the allowable type for SQLite but the main ones to get started with are:

+ NULL
+ Integer
+ Real
+ Text

To write a SQL scrpt, we need a text editor we can use to write our code. Any simple
code editor will do (atom, sublime, vscode,...). Go ahead an open up a blank document
in your text editor so it will be ready to go when we are ready to start writing some
code.

Go ahead and open your terminal and create a directory somewhere convenient
to store our example SQLite database and `.sql` scripts. 

`cd ~/Desktop`

`mkdir sqlite_exploration`

`cd sqlite_exploration`

We should now be ready to create a database!

### Thinking About Table Design - Schema

But what do we want to create?

Often I create small example tables that resemble data
that I need to work with but simplified. This way I can think through how to properly
store the data so that it is easy to update and report off of.

For example, in a past life I was a Lecturer at a university. If I wanted
to use SQL to manage and report on student grades, I might want to create a few
tables to help me manage the course data efficiently and set me up for efficient 
reporting later. 

+ eliminate duplication - We don't want to see the same information repeated
over and over in our tables. If an update is required, we want to update a single
table if possible.
+ operational efficiency - We should be able to easily query the data. Clear
and documented relationships between tables greatly simplifies the cognitive
overhead of keeping track of the data you are working with

With course grades as an example (I know, a little dull but at least most have
experience with courses and grading), I might create three tables to split the 
data I need to keep track of into
separate logical entities that have a clear relationship. 

Here is one possible simple schema for a gradebook:

STUDENTS - one row per student containing all student specific information: 

+ student_id - Integer Primary Key
+ student_name - Text
+ student_level (Freshmen, Sophmore, Junior, Senior) - Text
+ major - Text
+ email - Text

ASSIGNMENTS - one row per assignment containing all assignment specific information:

+ assignment_id - Integer Primary Key
+ assignment_category - Text
+ assignment_value - Real or Integer

GRADE - one row per graded assignment per student:

+ assignment_id - Integer
+ student_id - Integer
+ numeric_grade - Integer or Real

For the STUDENTS and
ASSIGNMENTS tables, we have a primary key (unique, non-null identifier) that we
can use to reference a specific student or assignment. The rest of the fields
in these two tables contain all the specific details about these entities. If
something needs to be updated, we are going to update it in one table. 

The final table, GRADE, is a *relation* table telling us how to join information
from the STUDENTS and ASSIGNMENTS tables. For example, we might want the email 
addresses of all students that scored below 75% on a particular assignment. Using
the GRADE table, SQL will be able to match rows using the `student_id` and 
`assignment_id` fields.

In SQLite, we use `CREATE TABLE tbl_name ();` with a comma separated list of columns
to add to our table along with information about type in the form `(name1 TYPE1, name2 TYPE2...)`. 

Go ahead and type the following
(or just copy it) into your text editor. Save the file with the name `table_schema.sql`.

```{r eval = FALSE, echo = TRUE }

CREATE TABLE students (
    student_id INTEGER PRIMARY KEY,
    name TEXT,
    student_level TEXT,
    major TEXT,
    email TEXT
);

CREATE TABLE assignments (
  assignment_id INTEGER PRIMARY KEY,
  assignment_category TEXT,
  assignment_value INTEGER
);

CREATE TABLE grade (
  assignment_id INTEGER,
  student_id INTEGER,
  numeric_grade REAL
  
)

```

To create a database with tables according to our schema, just type `$sqlite3 class.db < table_schema.sql`. Use `ls` to see that a new `class.db` file has been created in your
working directory.

To check everything is working, type `sqlite class.db` and a SQL prompt will appear. We
don't have any data in our tables yet, but you can type `.schema` to see a listing
of the tables we have just created.

```{r echo=TRUE, eval=FALSE}

(base) ➜  sql sqlite3 class.db                           
SQLite version 3.28.0 2019-04-15 14:49:49
Enter ".help" for usage hints.
sqlite> .schema
CREATE TABLE students (
    student_id INTEGER PRIMARY KEY,
    name TEXT,
    student_level TEXT,
    major TEXT,
    email TEXT
);
CREATE TABLE assignments (
  assignment_id INTEGER PRIMARY KEY,
  assignment_category TEXT,
  assignment_value INTEGER
);
CREATE TABLE grade (
  assignment_id INTEGER,
  student_id INTEGER,
  numeric_grade REAL
  
);
```

Type `.quit` to exit the SQL prompt and get back to the terminal.

### Adding Data to the Database Tables

Let's add some example data to our tables to play with. We are going to do it the
old fashioned way, insert rows into our tables...manually. We are keeping it simple
here, we can always add complexity later.

Open another blank document and name this file `class_data.sql`.

The simplest way for us to add data to our table is with 
`INSERT INTO tbl_name VALUES (val1, val2, ...)`. Notice in the code below, it is 
not required to list the column names in parentheses, but it can make it easier
to remember what goes where. 

You can just copy this piece into your script. For fun, add yourself to the student table,
add a new assignment to the assignments table and assign some random grades for your
new made up assignment.

Make sure to remember that primary keys must be unique, SQL will throw an error
if you try to add another observation with the same primary key.

```{r echo=TRUE, eval=FALSE}
/* STUDENTS */
INSERT INTO students (student_id, name, student_level, major, email)
  VALUES (1, 'Student 1', 'Freshman', 'CS', 'student1@uni.edu');

INSERT INTO students
  VALUES (2, 'Student 2', 'Freshman', 'CS', 'student2@uni.edu');
  
INSERT INTO students 
  VALUES (3, 'Student 3', 'Freshman', 'MATH', 'student3@uni.edu');
  
  
/* ASSIGNMENTS */
    
INSERT INTO assignments (assignment_id, assignment_category, assignment_value)
  VALUES (1, 'quiz', 10);
  
INSERT INTO assignments 
  VALUES (2, 'quiz', 10);
  
INSERT INTO assignments
  VALUES (3, 'exam', 100);
  
/* GRADES */
    
INSERT INTO grades (assignment_id, student_id, numeric_grade)
  VALUES (1, 1, 0.85);
  
INSERT INTO grades
  VALUES (2, 1, 0.75);
  
INSERT INTO grades
  VALUES (3, 1, 0.9);
  
INSERT INTO grades 
  VALUES (1, 2, 0.7);
  
INSERT INTO grades
  VALUES (2, 2, 0.6);
  
INSERT INTO grades
  VALUES (3, 2, 0.65);
  
INSERT INTO grades 
  VALUES (1, 3, 0.99);
  
INSERT INTO grades
  VALUES (2, 3, 1);
  
INSERT INTO grades
  VALUES (3, 3, .92);


```

Back at the terminal, type 
`sqlite3 -echo class.db < class_data.sql` 
to run the script `class_data.sql` on `class.db`. Now, the
tables have some data and we can start playing with SQL code.

## Read

To execute a simple query we can create a new script called `query.sql`.

The simplest type of query we might want to run, might look like the following:

```{r echo=TRUE, eval=FALSE}

SELECT (variables to select)
  FROM (table name to select from)
  WHERE (filtering conditions)

```

For example, add the following to your `query.sql` script.

```{r echo=TRUE, eval=FALSE}

SELECT *
  FROM students
  WHERE major = 'CS'

```

To run the script, type `sqlite3 -echo -column -header class.db < query.sql`. The
extra flags `-column -header` make the output much more readable.

```{r echo=TRUE, eval=FALSE}

$sqlite3 -echo -column -header class.db < scripts/query.sql

SELECT *
FROM students
WHERE major = 'CS'
student_id  name        student_level  major       email           
----------  ----------  -------------  ----------  ----------------
1           Student 1   Freshman       CS          student1@uni.edu
2           Student 2   Freshman       CS          student2@uni.edu
```


At this point, we are ready to start experimenting. For example, we could join all the tables together and select just the resulting columns of interest.

```{r echo=TRUE, eval=FALSE}

SELECT students.student_id,
       students.name,
       assignments.assignment_category,
       assignments.assignment_value,
       grades.numeric_grade
  FROM students, assignments, grades
  ON students.student_id = grades.student_id
    AND assignments.assignment_id = grades.assignment_id;
student_id  name        assignment_category  assignment_value  numeric_grade
----------  ----------  -------------------  ----------------  -------------
1           Student 1   quiz                 10                0.85         
1           Student 1   quiz                 10                0.75         
1           Student 1   exam                 100               0.9          
2           Student 2   quiz                 10                0.7          
2           Student 2   quiz                 10                0.6          
2           Student 2   exam                 100               0.65         
3           Student 3   quiz                 10                0.99         
3           Student 3   quiz                 10                1.0          
3           Student 3   exam                 100               0.92  

```


There are plenty of books, blogs, and websites to learn all about SQL syntax. One thing
that I find is missing at times is that hands on practice with tables that you created,
answering some questions that you might already know how to find using other means.

Recommended topics to explore:

+ the main sql verbs SELECT, WHERE, GROUP BY, HAVING, ORDER BY
+ joins: INNER JOIN, LEFT JOIN
+ aggregation functions: AVG(), COUNT(), SUM()



## Update

Updating tables is less common in data analysis and more common as part of 
a data pipeline, application, or other database process. Updates are typically not made willy-nilly, permanent changes to the data are made.

To UPDATE a value in a table, use the synatax

`UPDATE tbl_name SET var_to_change WHERE filter_for_where_to_change`

Here you can just create another new script `update.sql` and
copy in the following: 

```{r echo=TRUE, eval=FALSE}

UPDATE grades 
    set numeric_grade = 1.0 
    where student_id = 3 
        and assignment_id = 3;

```

Then run `sqlite3 class.db < update.sql`

and finally rerun our previous query to verify the values really did change

`sqlite3 -echo -column -header class.db < query.sql`

```{r echo=TRUE, eval=FALSE}
student_id  name        assignment_category  assignment_value  numeric_grade
----------  ----------  -------------------  ----------------  -------------
1           Student 1   quiz                 10                0.85         
1           Student 1   quiz                 10                0.75         
1           Student 1   exam                 100               0.9          
2           Student 2   quiz                 10                0.7          
2           Student 2   quiz                 10                0.6          
2           Student 2   exam                 100               0.65         
3           Student 3   quiz                 10                0.99         
3           Student 3   quiz                 10                1.0          
3           Student 3   exam                 100               1.0    
```

One dimension of richness of databases is the data integrity and data lineage
tooling that they provide. Here we are making simple updates but more generally,
in a business context, having precise control over what changes are made to the data
and an ability to track those changes is of utmost importance.

## Delete

### Tables

We can delete entire tables using `DROP TABLE tbl_name` but be careful, there is 
no recovery, If you have your table schema script separated like we did here, we can
recreate our database quickly.

Let's go ahead and add the following snippet at the top of our `table_schema.sql` script. 

```{r echo=TRUE, eval=FALSE}

DROP TABLE students;
DROP TABLE assignments;
DROP TABLE grades;

```

Aside from demonstrating how to delete a table, this will allow us run the script more than once without having
to manually `rm class.db` our database. Databases have built in mechanisms to prevent
tables from being overwritten which can be a pain when you are just playing around.

### Deleting Rows in Tables

Deleting an observation (or set of observations) is very similar to UPDATE; we
supply SQL with a WHERE condition helping us to identify which rows to remove.

Add the following code to your `update.sql` script.and rerun the script.

```{r echo=TRUE, eval=FALSE}

DELETE FROM assignments WHERE assignment_id = 3;

```

You should now see all the rows corresponding to `assignment_id = 3` have been
removed.

```{r echo=TRUE, eval=FALSE}

student_id  name        assignment_category  assignment_value  numeric_grade
----------  ----------  -------------------  ----------------  -------------
1           Student 1   quiz                 10                0.85         
1           Student 1   quiz                 10                0.75         
2           Student 2   quiz                 10                0.7          
2           Student 2   quiz                 10                0.6          
3           Student 3   quiz                 10                0.99         
3           Student 3   quiz                 10                1.0  

```


## Wrap Up

In a few short minutes, we were able to get SQLite3 up and running in the terminal,
we created a small example database and created small sql scripts to investigate basic SQL
CRUD operations. Of course there is much more to learn, but you now have the tools to play
around with schema design and query writing...all straight from the terminal.


In a future post, we will continue on with writing some more involved SQL report writing
techniques and incorporate proper databases into our R, python, or SAS projects.














